Psyc 601 Multivariate Statistics Project

### 2013-08-23 writeups/proposal/proposal.tex
Fri commits
Former-commit-id: 5728261a08d655cee7c1f1ed3c06b1bc541ba922 [formerly d0e58081e5d89f1524b1f0259fc8d235b12a2809]
Former-commit-id: be46e14b71525979d6101ae9753ca692447f7678

\usepackage{textcomp}
\usepackage[american]{babel}
\usepackage[utf8]{inputenc}

@@

\usepackage{amssymb,amsmath}
\usepackage{setspace}

@@

\cite{Anderson2004}

@@

>>>>

\usepackage{xfrac}
\usepackage{textcomp}
\usepackage[american]{babel}
\usepackage[utf8]{inputenc}

@@

\usepackage{tabu}
\usepackage{amssymb,amsmath}
\usepackage{setspace}

@@

ACT-R \cite{Anderson2004} is a cognitive architecture that formalizes how each cognitive process of the brain (e.g., memory, learning, visual and motor) interacts to produce behavior.
The declarative memory system is a component of that architecture that models the timing, learning, and forgetting processes that make up long-term declarative memory retrieval.
The equations that make up this system can be described through a rational analysis of long-term memory retrieval.
That is, given the task of retrieving a chunk of information from long-term declarative memory,
the current context (i.e., external and internal environment state), and past experience (i.e., prior memories and exposure), 
what is the optimal behavior (i.e., the optimal chunk to retrieve from memory)?
Using Bayesian reasoning, each chunk of information in declarative memory can be assigned a prior likilihood of being retrieved again, given the prior number of times it has been needed.
Those prior probabilities for each chunk are then adjusted for the current context, so that the posterior probabilities represent the likilihood that a chunk is needed, given prior odds and current environment state.

@@

\begin{table}[!ht]
  \caption{ACT-R Declarative Memory Model}
  \label{tabACT-RModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Activation &	 	$A_{i} = B_{i} + \sum_{j \in c}^{} W_{j} S_{ji}$ \\
      Attentional Weight &	$W_{j} = \frac{W}{n}$ \\
      Base Level & 		$B_{i} = log \sum_{j=1}^{n} {t_{j}}^{-d}$ \\
      Constant Base Level &	$B_{i} = log \frac{p(i)}{p(\overline{i})}$ \\
      Strength of Association &	$S_{ji} = log \frac{p(i|j)}{p(i|\overline{j})} \approx log \frac{p(i|j)}{p(i)}$ \\
      Recall Probability &	$P_{i} = \left( 1 + e^{\frac{\tau - A_{i}}{s}} \right )^{-1}$ \\
      \hline
    \end{tabu}
  }
\end{table}

A formal description of the ACT-R Declarative Memory model is included in Table \ref{tabACT-RModel}.
The total activation ($A_{i}$) for a chunk in declarative memory is a function of two components: base level activation ($B_{i}$) and strength of association ($S_{ji}$).
The recall probability that a chunk will be retrieved from memory increases with total activation ($A_{i}$).

\subsubsection{Base-Level Activation}

Base-level activation reflects the log prior odds of needing an observed chunk again.
The primary way to calculate these log prior odds is to use the standard base level equation in Table \ref{tabACT-RModel}.
This equation formalizes how log prior odds are a function of both frequency and recency of prior exposure to a particular chunk.
Chunks used more frequently (either through exposure or from a retrieval) are more likely to be needed for retrieval again.
However, as time progresses and a particular chunk is no longer used, the activation for that chunk decays.
In this way the standard base level equation formalizes the time dynamics of the retrieval system, where a chunk's base-level activation evolves over time, depending on its current frequency and recency of use.

For some domains, it is reasonable to assume that the base-level activations of each chunk within a particualar window of interest are stable, and do not change within that window.
Programming-language popularity over the past few years is a reasonable example of this.
Although the popularity of various programming languages has certainly changed slightly over the past few years (.e.g, \emph{closure}'s growth), it is not the case that the changes have been drastic.
\emph{C\#}, \emph{Python}, and \emph{Java} are still a few of the most popular languages, \emph{Common Lisp} has not gained or lost much ground in popularity over the past few years, etc.

Further, if this is a reasonable assumption to make, it greatly simplifies the computation of the base-level activation for chunks, and can turn the computation into a tractable problem for large datasets.
In these time-constant domains, one can compute the log prior odds of needing each chunk directly as the log odds ratio of being exposed to a chunk.
For example, if the StackOverflow tag \emph{PHP} has been used four times as often as the tag \emph{Common Lisp}, then the prior odds of needing \emph{PHP} again is $\frac{.8}{.2}=4$ times that of \emph{Common Lisp}.

\subsubsection{Strength of Association}

Strength of association ($S_{ji}$) reflects the amount of log odds adjustment to the activation of a chunk, given the current context (i.e., external environment and internal state).
Context for the StackOverflow domain for example, can be represented as each word in the title and body of a post.
Context for the Twitter domain would be the words in a Tweet.
Association strenght between a chunk in memory and a single contextual element can be computed directly by calculating its context-adjusted odds ratio:
The likelihood a chunk occurred with the current context ($p(i|j)$) over the likeihood that the chunk occured in any of the other contexts ($p(i|\overline{j})$).
For large datasets, the likelihood that a chunk occurs in any particular context reaches near-zero values ($p(j) \Rightarrow 0$).
So it can be assumed that the likelihood that a chunk occurs in any context but one ($p(i|\overline{j})$) is equivalent to the likelihood that a chunk occurs in any context ($p(i)$).

Using this assumption for large datasets, the interpretation of the context-adusted odds ratio becomes much simpler.
If this ratio for a particular tag and context chunk is greater than one, then the log is positive, which means that this context has been observed more often with this tag than in general.
If this ratio is less than one, then the log is negative, which provides a negative adjustment of total activation since this context has been observed more often in general than with this particular tag.
